---
title: "Temperature"
output: html_document
---

```{r setup, include=FALSE}
library(tidyverse)
library(janitor)
library(broom)
library(stringr)
library(leaflet)
library(leaflet.extras)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
knitr::opts_chunk$set(echo = TRUE)
```

# NYC Environment and Health Data

## About the Dataset

This analysis includes datasets that were retrieved from the number of  [Climate Datasets](https://a816-dohbesp.nyc.gov/IndicatorPublic/beta/data-explorer/climate/?id=2141#display=summary) publicly available by NYC.gov. The datasets used are "Daytime Summer Surface Temperatures" and "Heat Vulnerability Index (NTA)". The NTA codes were used to merge the two datasets.

```{r, echo = FALSE, include = FALSE}
envo_health = read_csv("data/envo_health.csv") %>% 
  clean_names()

envo_health = 
  envo_health %>% 
  mutate(
    daytime_surface_temp = (daytime_surface_temp- 32)*.5556 
  )
```

# Summary Statistics
```{r, echo = FALSE}
envo_health %>% 
  group_by(borough) %>% 
  summarise(
    mean_daytime = mean(daytime_surface_temp)
  ) %>% 
  knitr::kable(digits = 3, align = "cc", caption = "Table 1: Mean Surface Daytime Temperature by Borough", col.names = c("Borough", "Mean Temperature (°C)"))

envo_health %>% 
  group_by(borough) %>% 
  summarise(
    mean_daytime = mean(heat_vulnerability_index)
  ) %>% 
  knitr::kable(digits = 3, align = "cc", caption = "Table 2: Mean Heat Vulnerability Index by Borough", col.names = c("Borough", "Mean Heat Vulnerability Score"))
```
Score

Each neighborhood receives a score from 1 (lowest risk) to 5 (highest risk) based on a statistical model that uses social and environmental factors to estimate risk of heat-related death across NYC neighborhoods. A low vulnerability score does not mean no risk. Every neighborhood has residents at risk for heat illness and death. Residents who are at highest heat risk are those who do not have or use air conditioning and are older or have chronic health conditions or serious mental illnesses

# New York City Map
#### Daytime Surface Temperature and Heat Vulnerability Index by Location
```{r, echo = FALSE}
pal <- colorNumeric(
  palette = "viridis",
  domain = envo_health$daytime_surface_temp)
envo_map = envo_health %>% 
  mutate(
    daytime_surface_temp = round(daytime_surface_temp, digit = 2),
    click_label = 
      str_c("<b>Temperature (°C): ", daytime_surface_temp, " <br> Heat Vulnerability Score: ", heat_vulnerability_index, " <br> Neighborhood: ", neighborhood)) %>% 
    leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(lng = ~longitude, lat = ~latitude, radius = .1, color = ~pal(daytime_surface_temp), popup = ~click_label) %>% 
  setView(lng=-73.935242, lat=40.730610, zoom = 11) %>% 
  addLegend("bottomright", pal = pal, values = ~daytime_surface_temp,
    title = "Temperature",
    labFormat = labelFormat(suffix = "°C"),
    opacity = 1
  )
  
envo_map
```

# Univariate Models

Two univariate, linear models were built to explore the association between:

  - boroughs and daytime surface temperature
  - boroughs and heat vulnerability index

Univariate models were because only one predictor was used. Linear models were used because the outcome variable, `daytime_surface_temp`, was continuous.

#### Borough and Daytime Surface Temperature  

In the summary statistics, the mean daytime surface temperature was not appreciably different across the boroughs. A linear model was used to further explore if there was a relationship between borough and daytime surface temperature.
```{r, echo = FALSE}
envo_health = 
  envo_health %>% 
  mutate(
    borough = fct_relevel(borough, "Manhattan"))

envo_surfmodel =
  lm(daytime_surface_temp ~ borough, data = envo_health) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

envo_surfmodel %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 4: Coefficient Estimates of Linear Model Exploring Association of Borough and Daytime Surface Temperature", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```


#### Borough and Heat Vulnerability Index

In the summary statistics, Bronx had the highest mean heat vulnerability score, which makes it the most at-risk borough in NYC to heat-related deaths. A linear model was used to further explore the relationship between borough and heat vulnerability index.

```{r, echo = FALSE}
envo_heatmodel =
  lm(heat_vulnerability_index ~ borough, data = envo_health) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

envo_heatmodel %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 5: Coefficient Estimates of Linear Model Exploring Association of Borough and Heat Vulnerability Index", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```



# Urban Green Spaces, Daytime Surface Temperature, and Heat Vulnerability Index in New York City
```{r, echo = FALSE, include = FALSE}
UG_df = read_csv("./data/GreenRoofData2016_20180917.csv", na = c("","Unknown")) %>% 
  janitor::clean_names() %>% 
  select(fid:notes, address:ycoord, -doitt_id, -qa, -feat_code, -spdist1, -bbl_fixed) %>% 
  rename(
    UG_id = fid,
    latitude = ycoord,
    longitude = xcoord,
    cons_year = cnstrct_yr,
    green_area = gr_area,
    building_area = bldg_area
  ) %>% 
  mutate(
    prop_gr = prop_gr*100,
    borough = recode(borough,
         BK = "Brooklyn",
         BX = "Bronx",
         MN = "Manhattan", 
         QN = "Queens",
         SI = "Staten Island"),
    ownertype = recode(ownertype,
         C = "Public",
         M = "Other",
         O = "Public", 
         P = "Private",
         X = "Other",
         P = "Private"),
    ownertype = replace_na(ownertype, "Private"),
    green_area = green_area*0.09290304,
    building_area = building_area*0.09290304,
    heightroof = heightroof*0.3048,
    groundelev = groundelev*0.3048
  ) %>% 
  select(UG_id, cons_year, borough, zonedist1, address, ownertype, longitude, latitude, green_area:groundelev, everything()) %>% 
  drop_na(green_area) 
```

```{r, echo = FALSE}
pal <- colorNumeric(
  palette = "viridis",
  domain = envo_health$daytime_surface_temp)
envoroof_map = envo_health %>% 
  mutate(
    daytime_surface_temp = round(daytime_surface_temp, digit = 2),
    click_label = 
      str_c("<b>Temperature (°C): ", daytime_surface_temp, " <br> Heat Vulnerability Score: ", heat_vulnerability_index, " <br> Neighborhood: ", neighborhood)) %>% 
    leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(lng = ~longitude, lat = ~latitude, radius = 2, color = ~pal(daytime_surface_temp), popup = ~click_label) %>% 
  addMarkers(data = UG_df, ~longitude, ~latitude, icon = list(iconUrl = 'https://icons.iconarchive.com/icons/icons8/christmas-flat-color/512/christmas-tree-icon.png',  iconSize = c(5, 10))) %>% 
  setView(lng=-73.935242, lat=40.730610, zoom = 11) %>% 
  addLegend("bottomright", pal = pal, values = ~daytime_surface_temp,
    title = "Temperature",
    labFormat = labelFormat(suffix = "°C"),
    opacity = 1
  ) 

envoroof_map
```
**The Christmas trees in the urban green spaces, daytime surface temperature, and heat vulnerability index New York City map represent the urban green spaces in New York City.**

```{r, echo = FALSE, include = FALSE}
envo_greenroof = read_csv("data/envo_greenroof.csv") %>% 
  clean_names()

envo_greenroof = 
  envo_greenroof %>% 
  mutate(
    borough = fct_relevel(borough, "Manhattan"),
    avg_daytime_temp = (avg_daytime_temp- 32)*.5556) 
```

```{r, echo = FALSE}
envo_greenroof_surfmodel = 
  lm(avg_daytime_temp ~ avg_green_area, data = envo_greenroof) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

envo_greenroof_surfmodel %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 6: Coefficient Estimates of Linear Model Assessing the Association Between Average Green Roof Surface Area and Average Daytime Surface Temperature", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))

envo_greenroof_heatmodel = 
  lm(avg_heat_index ~ avg_green_area, data = envo_greenroof) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

envo_greenroof_heatmodel %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 7: Coefficient Estimates of Linear Model Assessing the Association Between Average Green Roof Surface Area and Average Heat Vulnerability Index", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```

# Limitations

There are limitations to the temperature analyses. 

1. The datasets vary in the date of collection. Although the `envo_health` dataset is comparable by neighborhood, the daytime surface temperature data were collected on a single day and the heat vulnerability index data were collected across a few years. 

2. The temperature datasets and the primary dataset are not collected from the same location. It may be interesting to explore the association of green space are and temperature predictors from data that were collected at the same coordinate. This may give a better understanding of the impact of green space area and at risk neighborhoods. 