---
title: 'Urban Greenspaces: Exploratory Analysis'
author: "mm5951"
date: "`r Sys.Date()`"
output: html_document
---

**PENDING** : assess which libraries to include.

```{r, include = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(leaflet)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

The primary goal of this exploratory data analysis (EDA) is to provide a primary dataset of NYC Urban Greenspaces upon which regression analysis can be conducted with potentially associated dimensions (temperature, econometrics, and health outcomes).

Secondly, this EDA aims to provide:

* a descriptive analysis of the NYC Urban Greenspaces primary dataset;
* relevant data visualizations of the above (notably including a map using the `leaflet` package and overtime trends);
* **TBD** a test of homogeneity to assess the distribution of Urban Greenspaces (UG) at the City Council District and Borough levels.

## Data load & wrangling

```{r, message = FALSE, warning = FALSE}
UG_df_raw = read_csv("./data/GreenRoofData2016_20180917.csv", na = c("","Unknown")) 

# Variables contained in the dataframe
str(UG_df_raw)

# Quick summary
skimr::skim(UG_df_raw)

# Looking into key variables
UG_df_raw %>%
  count(qa)

UG_df_raw %>%
  count(borough)
```

## Descriptive analysis

```{r}
UG_df = read_csv("./data/GreenRoofData2016_20180917.csv", na = c("","Unknown"))
```

## Data visualization

Using the `leaflet` package to plot observations on a NY map.

Overtime trends (# of UG inaugurated per year).

## Test of homogeneity



