---
title: "Urban Green Spaces in NYC"
output: html_document
---

```{r, include = FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(broom)
library(janitor)
library(leaflet)

knitr::opts_chunk$set(
	echo = TRUE,
	warning = FALSE,
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))
options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
```

Roofs planted with vegetation — **Urban Green Spaces (UGS)** — can help cities adapt to a changing climate by absorbing storm water, lowering local temperatures, and providing insulation that cuts indoor heating and cooling costs. Yet green roofs **cover less than 0.1% of New York City’s 1 million buildings**, according to a [new analysis performed by colleagues at Columbia University](https://news.climate.columbia.edu/2022/11/01/new-study-provides-first-look-at-green-roof-distribution-across-nyc/).

Hereby we present a **secondary analysis of the publicly available dataset**, which includes data from n=736 UGS as of 2016. In complementary analyses found within the project website, we **draw links to economical, clinical and environmental factors** which could potentially be associated with the distribution of UGS within New York City (NYC).

## Goals of the analysis

The primary goal of this analysis is to provide a primary dataset of NYC Urban Green Spaces upon which regression analysis can be conducted with potentially associated dimensions (temperature, econometrics, and health outcomes).

Secondly, this analysis aims to provide:

* a descriptive analysis of the NYC Urban Green Spaces (UGS) primary dataset, as well as stratified analysis at the City Council District and Borough levels;
* relevant data visualizations of the above (notably including a map using the `leaflet` package);
* an attempt to fit the dataset into a linear regression model to elucidate differences in the distributuion of UGS surfaces at the City Council District and Borough levels.

## Data load & wrangling

The **primary Urban Green Spaces (UGS) dataset** is loaded and wrangled to generate a **`UG_df` dataframe**. The dataset is available in the "data" folder within the project's repository. The following procedure is then applied: 

* select relevant variables using `select()`
* rename certain variables for comprehension using `remame()`
* re-order the variables using `arrange()`
* drop entries with now total green area available using `drop_na()` 
* calculate percentage of green coverage of each building, as well as `recode()` borough names and ownership status, and convert sqm int sqm, using `mutate()`
* note ownertypes are aggregated into "Public", "Private" and "Other" as per mapPLUTO definitions (refer to the codebook for more details). Given the presence of "na", the `replace_na()` function is employed.

```{r, message = FALSE, warning = FALSE}
UG_df = read_csv("./data/GreenRoofData2016_20180917.csv", na = c("","Unknown")) %>% 
  janitor::clean_names() %>% 
  select(fid:notes, address:ycoord, -doitt_id, -qa, -feat_code, -spdist1, -bbl_fixed) %>% 
  rename(
    UG_id = fid,
    latitude = ycoord,
    longitude = xcoord,
    cons_year = cnstrct_yr,
    green_area = gr_area,
    building_area = bldg_area
  ) %>% 
  mutate(
    prop_gr = prop_gr*100,
    borough = recode(borough,
         BK = "Brooklyn",
         BX = "Bronx",
         MN = "Manhattan", 
         QN = "Queens",
         SI = "Staten Island"),
    ownertype = recode(ownertype,
         C = "Public",
         M = "Other",
         O = "Public", 
         P = "Private",
         X = "Other",
         P = "Private"),
    ownertype = replace_na(ownertype, "Private"),
    green_area = green_area*0.09290304,
    building_area = building_area*0.09290304,
    heightroof = heightroof*0.3048,
    groundelev = groundelev*0.3048
  ) %>% 
  select(UG_id, cons_year, borough, zonedist1, address, ownertype, longitude, latitude, green_area:groundelev, everything()) %>% 
  drop_na(green_area) 
```

Notably, the same authors provide an aggregated dataset at the City Council District which is loaded as a **secondary UGS dataset** under the **`UG_agg_df`** name. A similar procedure as described above is used.

```{r, message = FALSE, warning = FALSE}
UG_agg_df= read_csv("./data/NYC_CounDist_GreenRoof_Bldg_MaxHVI_PctCSO.csv", na = c("","Unknown")) %>% 
  janitor::clean_names() %>% 
  select(coun_dist:prop_greenroofarea_of_bldgarea) %>% 
  rename(
    green_area = greenroof_area_sqmeters,
    building_area = bldg_area_total_sqmeters,
    prop_gbuild = prop_greenroofs_of_bldgs,
    prop_gr = prop_greenroofarea_of_bldgarea
  ) %>% 
  mutate(
    prop_gr = prop_gr*100,
    prop_gbuild = prop_gbuild*100,
    green_area = green_area*0.09290304,
    building_area = building_area*0.09290304
  )
```

Note that all variable names are available in the [project's codebook](codebook.html).

## Descriptive analysis

Next, we perform a descriptive analysis of the NYC Urban Green Spaces (UGS) primary dataset, as well as stratified analysis at the City Council District and Borough levels.

### Overall `UG_df` Datset Descriptive Analysis

Overall, the "UG_df" contains contains `r ncol(UG_df)` variables related to `r nrow(UG_df)` urban green spaces in NYC. The total UGS area in NYC sums to `r sum(UG_df$green_area)` square meter. The following table summarizes the average UGS in NYC, including its size (in square meter), height (in meter) and percentage of green coverage (that is, the proportion of green space within the total building area). Key variables are summarized in the table below using `knitr:kable()`.

```{r}
UG_df %>% 
  summarize(
    mean_green = mean(green_area, na.rm = TRUE),
    median_green = median(green_area),
    min_green = min(green_area),
    max_green = max(green_area),
    mean_height = mean(heightroof, na.rm = TRUE),
    perc_green = mean(prop_gr, na.rm = TRUE)
    ) %>% 
    knitr::kable(digits = 3, align = "lcccc", caption = "Table 1: Summary Metrics of Urban Green Spaces (UGS) in New York City (NYC), including surface metrics (in sqm, mean, median and range), building height (in m) and proportion of UGS coverage from total building surface (%)", col.names = c("Mean surface", "Median surf.", "Minimum surf.", "Maximum surf.", "Mean building height", "Percent of UGS"))
```

Overall, within New York City, UGS have an **mean size of 334.52 sqm (range 0.94-25,763.66 sqm)**. The broad range of values, as well as the median situated at 109.06 sqm (way below the mean), it suggests this variable's **right-skewed distribution**. We will explore this more in the "data visualization" section below. On average, UGS in NYC are situated at **43.47 m of height**, and **cover 19.63% of the total building's surface**.

### Stratified Analysis by Borough

When stratified by borough, differences in the number and dimensions of UGS become apparent. Following a similar procedure as above, we generate a summary table below, arranged according to the amount of UGS present in each borough.

```{r}
UG_df %>% 
  group_by(borough) %>% 
  summarize(
    n = n(),
    perc = n()/736*100,
    tot_green = sum(green_area),
    mean_green = mean(green_area, na.rm = TRUE),
    median_green = median(green_area),
    min_green = min(green_area),
    max_green = max(green_area),
    mean_height = mean(heightroof, na.rm = TRUE),
    perc_green = mean(prop_gr, na.rm = TRUE)
    ) %>% 
  arrange(desc(n)) %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 2: Summary Metrics of Urban Green Spaces (UGS) in New York City (NYC) by Borough")
```

**Manhattan** contains over 63% of all NYC's UGS (n=465), situated at a considerably higher altitude (mean height 54.24 m). Nevertheless, its UGS are also the smallest in percentual terms when asessing the proportion of building coverage (15.24%). Conversely, **Queens** and **Staten Island** have the lowest amoung of UGS (n=50 and n=4, respectively), but their UGS area the biggest on average (mean sizes of 468.55 and 456.72 sqm, respectively). These results could be linked to the property prices in each areas (further analysis would be needed to assess so).


### Stratified Analysis by City Council District 

Although NYC city council districts typically correspond to borough sub-divisions, the equivalence is not perfect (note council district #8 belongs to both Manhattan and Bronx, more info [here](https://council.nyc.gov/districts/)). In that sense, the same authors provide a similar dataset aggregated at the City Council District level, which we have loaded with the `UG_agg_df` name. A similar descriptive analysis as above is performed. Notably, in this dataset the total number of buildings per district is available, and therefore we compute the percentage of buildings with an UGS.

```{r}
UG_agg_df %>% 
  group_by(coun_dist) %>% 
  summarize(
    borough = borough,
    n = num_greenroofs,
    perc_UGS = num_greenroofs/736*100,
    perc_building = num_greenroofs/num_bldgs_total*100,
    mean_green = mean(green_area/num_greenroofs, na.rm = TRUE),
    ) %>% 
  arrange(desc(n)) %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 3: Summary Metrics of Urban Green Spaces (UGS) in New York City (NYC) by City Council District")
```

As a result, we observe that **3 City Council Districts in Manhattan concentrate over half of the borough's UGS (n=318)**, with over 1% of its buildings covered by green spaces. Conversely, **8 districts within Bronx, Queens, Brooklyn and Staten Island have no UGS at all**. These results suggest the existence of distribution differences, which will be addressed in the last section of this analysis.



## Data visualization

### Mapping UGS in NYC
Using the `leaflet` package, we plot each UGS on a NYC map. A `pal()` function is generated to establish a color according to the size of the UGS.

```{r}
pal <- colorNumeric(
  palette = "viridis",
  domain = UG_df$green_area)

UG_map = UG_df %>% 
  mutate(
    green_area = round(green_area, digit = 2),
    heightroof = round(heightroof, digit = 2),
    click_label = 
      str_c("<b>Area ", green_area, " sqm</b><br>Height ", heightroof, " m<br>")) %>% 
  leaflet() %>% 
  addProviderTiles(providers$CartoDB.Positron) %>% 
  addCircleMarkers(~longitude, ~latitude, radius = .1, color = ~pal(green_area), popup = ~click_label)

UG_map
```

As described above, we observe a concentration of UGS in southern Manhattan. Scarce green spaces are present in southern Brooklyn, Staten Island, Queens or the Bronx.


### Distribution of UGS surface by borough

Next, the distribution of UGS surface by borough is investigated. The `UG_df` is first plotted on a boxplot to have a better understanding of outliers (as we have indication of a right-skewed disribution according to the descriptive analysis above). 

```{r}
UG_boxplot = UG_df %>% 
  ggplot(aes(x = green_area, y = borough)) + 
  geom_boxplot() +
  labs(
    title = "Distribution of UGS surface by borough",
    x = "Green Area (sqm)",
    y = "Borough")

UG_boxplot
```

Given the presence of considerable outliers, the plotted green area surface is then limited to 10,000 sqm and visualized through a violin plot.

```{r}
UG_violin = UG_df %>% 
  filter(green_area <10000) %>% 
  ggplot(aes(x = green_area, y = borough)) + 
  geom_violin(aes(fill = borough), color = "blue", alpha = .5) +
  theme(legend.position = "bottom") +
  labs(
    title = "Distribution of UGS surface by borough",
    x = "Green Area (sqm)",
    y = "Borough")

UG_violin
```

Overall, these visualizations confirm the sparse distribution of UGS surface area (sqm) accross boroughs, as well as suggest differences according to its location (see map above). We will investigate this further through regression in the upcoming section.

### UGS development overtime in NYC

Finally, the construction of UGS trends is investigated overtime. To do so, the `cumsum()` function is used over a generated dummy variable that allocates the value "1" to all UGS. Note that the observations with an invalid or missing construction year are omited.

```{r}
UG_overtime = UG_df %>%
  drop_na(cons_year) %>% 
  filter(cons_year !=0) %>% 
  arrange(cons_year) %>% 
  mutate(
    cumsum_dummy = 1,
    cumulative_UG = cumsum(cumsum_dummy)
  )
  
UG_overtime %>% 
  group_by(borough) %>% 
  ggplot(aes(cons_year, cumulative_UG, color = borough)) +
  geom_point(alpha = .5) +
  labs(
    title = "Overtime construction of Urban Green Spaces (UGS)",
    x = "Time (year)",
    y = "Number of UGS")
```

We observe a progressive construction of UGS overtime, with a rampant trend as of early 2000s (potentially linked to urban planning regulations and greater interest in climate change mitigation strategies).

## Regression analyses

In this section, we perform different regression analyses to understand whether the location within NYC (x: predictor variables, both "borough" and "city council district") influences either the size or number of UGS (y: outcome variable, both "green_area" and "num_greenroofs").

### UGS number by borough and city council district
We first investigate whether there are differences in the amount of UGS by borough and city council district (using the `UG_agg_df`). We run linear regression `lm()` models.

```{r}
NumberUG_borough_model = 
  lm(num_greenroofs ~ borough, data = UG_agg_df) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

NumberUG_borough_model %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 4: Coefficient Estimates of Linear Model Assessing the Association Between Number of Green Roof Spaces and Borough in NYC", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```

```{r}
NumberUG_CounDist_model = 
  lm(num_greenroofs ~ coun_dist, data = UG_agg_df) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

NumberUG_CounDist_model %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 5: Coefficient Estimates of Linear Model Assessing the Association Between Number of Green Roof Spaces and City Concil District in NYC", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```

### UGS surface by borough 
Next, we investigate differences on the UGS surface area by borough. Given the presence of outliers and a right-skewed distribution observed in the descriptive analysis above, we apply a Kruskal-Wallis test.

```{r}
UGarea_borough_model = 
  lm(green_area ~ borough, data = UG_df) %>% 
  tidy() %>% 
  select(-statistic) %>% 
  mutate(
    low_conf = estimate - 1.96*std.error,
    upper_conf = estimate + 1.96*std.error) %>%  
  select(-std.error)

UGarea_borough_model %>% 
  knitr::kable(digits = 3, align = "lcccc", caption = "Table 6: Coefficient Estimates of Linear Model Assessing the Association Between Green Area of Green Roof Spaces and Borough in NYC", col.names = c("Term", "Estimate", "p-value", "Lower Confidence Interval", "Upper Confidence Interval"))
```

```{r}
fit3 = kruskal.test(green_area ~ borough, data = UG_df)

fit3 %>% 
  broom::tidy()
```

**Interpretations of the lm models pending**